---
title: "Finish Aligning and Generate Read Counts"
author: "Jiayi Fan"
date: "2025-03-31"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_depth: 2
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE)
```

The goal of this week’s homework is to complete the alignment and read
count generation of all the samples of your project data. You all had a
first go at it during the HW of week 6. Now is the time to get back to
that, change what needs to be changed and write a script that will
perform the alignment for all your samples. The final step for this week
should generate a table of read counts which you should read into R.

1.  Set up a github repository where you will store all your scripts,
    and your final report. Submit the URL of your repo. If you want to
    set it to private, you will have to grant us access, i.e., Luce
    (lskrbnek), Merv (mfansler), and Karolina (sienkie).
2.  Align all your samples. Ideally use a for-loop within a script,
    i.e., automate and standardize the task to a certain extent, but do
    remember that legibility is valuable, too.
3.  Generate a read count table.\*
4.  Load the read count table into R and perform the quality controls
    and processing steps that we discussed in class.

The github link for this homework is https://github.com/jiayifan-jf923/angsd-hw.

## Load Data and Normalization

```{r library__data}
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(pheatmap))

# Set folder and read in featureCounts output
folder <- "gene_counts/"  # <--- change this to your actual path
counts_file <- paste0(folder, "fc_output")  # adjust filename if needed

df_counts <- read.table(counts_file, header = TRUE, comment.char = "#", check.names = FALSE)
head(df_counts)
```

Here I loaded the read counts as tables, filter out genes with zero total counts, and normalize for sequencing depth.

```{r normalize}
# Extract SRR id
colnames(df_counts) <- gsub(".*/(SRR[0-9]+).*", "\\1", colnames(df_counts))

srr_ids <- colnames(df_counts)[-(1:6)]
if (length(srr_ids) != 6) {
  stop("Expected 6 samples but found ", length(srr_ids))
}

rownames(df_counts) <- make.names(df_counts$Geneid, unique = TRUE)
cts_gene_sample <- df_counts[, srr_ids]

sample_conditions <- c("shWT1", "shWT1", "shWT1_dox", "shWT1_dox", "LP9_ctrl", "LP9_ctrl")

df_coldata <- data.frame(
  condition = factor(sample_conditions),
  row.names = srr_ids
)

# Create DESeqDataSet
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = cts_gene_sample,
  colData = df_coldata,
  design = ~ condition
)

rowData(dds) <- df_counts[, 1:6]

# Filter out genes with zero total counts
dds <- dds[rowSums(counts(dds)) > 0, ]

# Normalize for sequencing depth
dds <- estimateSizeFactors(dds)
print(sizeFactors(dds))
```


```{r qc}
# QC plot: raw vs normalized counts
par(mfrow = c(1, 2))
plot(density(log2(counts(dds)[,1] + 1)), col = 1, main = "Raw counts (log2)", xlab = "log2(counts + 1)")
for (i in 2:ncol(dds)) {
  lines(density(log2(counts(dds)[,i] + 1)), col = i)
}

plot(density(log2(counts(dds, normalized = TRUE)[,1] + 1)), col = 1, main = "Normalized counts (log2)", xlab = "log2(norm counts + 1)")
for (i in 2:ncol(dds)) {
  lines(density(log2(counts(dds, normalized = TRUE)[,i] + 1)), col = i)
}
```

```{r}
raw_totals <- colSums(counts(dds))
norm_totals <- colSums(counts(dds, normalized = TRUE))

barplot(rbind(raw_totals, norm_totals), beside = TRUE,
        names.arg = colnames(dds), legend.text = c("Raw", "Normalized"),
        main = "Library size before vs after normalization", las = 2)
```
After normalization, the read count across samples are the same, so the size factor normalization worked as expected.

## Visualization (correlation heatmaps, PCA)

```{r cluster}
# Log transformation
rlog_dds <- rlog(dds)

# Compute distance matrix
corr_matrix <- cor(assay(rlog_dds), method = "pearson")
pheatmap(1 - corr_matrix, main = "1 - Pearson Correlation")

# Hierarchical clustering
as.dist(1 - corr_matrix) %>%
  hclust() %>%
  plot(labels = colnames(corr_matrix), main = "Sample Clustering (rlog)")

# PCA plot
plotPCA(rlog_dds, intgroup = "condition") +
  ggtitle("PCA of Samples")
```

According to the PCA and clustering, there are clear separation of different conditions, and similarity within same conditions.

## Differential Expression Analysis

```{r deg}
# Ensure WT is reference level
dds$condition <- relevel(dds$condition, ref = "LP9_ctrl")

# Run DESeq2 analysis
dds <- DESeq(dds)

# Get results with adjusted p-values (FDR)
res <- results(dds, independentFiltering = TRUE, alpha = 0.05)

# Summary of results
summary(res)

# Histogram of adjusted p-values
hist(res$padj, breaks = 40, main = "Adjusted p-values")

# Sort by significance
res_sorted <- res[order(res$padj), ]
head(res_sorted)

```

```{r visualize_deg}
# MA plot (log2FC vs mean normalized counts)
plotMA(res, alpha = 0.05, ylim = c(-4, 4), main = "MA Plot")

# Volcano plot (requires EnhancedVolcano)
# BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)
EnhancedVolcano(res,
    lab = rownames(res),
    x = "log2FoldChange",
    y = "padj",
    title = "Volcano Plot: Tumor vs Control")

```

Select genes with adjusted p-value < 0.05 for heatmap. The ploted heatmap shows z-score-scaled expression across samples.

```{r}
library(pheatmap)

# Subset genes
sig_genes <- rownames(subset(res, padj < 0.05))
rlog_mat <- assay(rlog_dds)[sig_genes, ]

pheatmap(rlog_mat, scale = "row", show_rownames = FALSE,
         main = "DEGs (padj < 0.05, z-score scaled)")

```




